
COMP356 Technical Report 2 Code

Original file is located at
    https://colab.research.google.com/drive/10Cq8ymgdLykexDemWkhfl_ZkHHBGWRPl

# Technical Report 2

This is where we wrtie code and run.
"""
```{r}
library(Matrix)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tibble)
library(janitor)
source('https://www.macalester.edu/~dshuman1/365/365Functions.r')
```

### Import Data
We might need to **upload the file** to the localstorage on this Colab every session.

```{r}


#data frame 
data <- as.data.frame(read.csv('nhl2122.csv')) %>%
    clean_names() %>% 
    rename(visitor_score = g, home_score = g_1) %>% 
    select(-c(x, att, log, notes, date))
# team vector for index sorted in aphabetically order
teams <- sort(union(unique(data$visitor), unique(data$home)))

head(data)
```

### Data Summary

```{r}
data_summary = data %>% summarise(
    win = ifelse(visitor_score > home_score, visitor, home), 
    lose = ifelse(visitor_score < home_score, visitor, home), 
    score = abs(visitor_score - home_score))%>% 
    pivot_longer(cols=c(win, lose), values_to = 'names', names_to = 'result') %>%
    mutate(result = as.factor(result))%>% 
    group_by(names, result) %>% 
    summarise(diff_by_total = sum(score), count = n()) %>% 
    pivot_wider(
        id_cols = names,
     names_from = result,
  values_from = c(diff_by_total, count)) %>% 
  mutate( net_score = diff_by_total_win - diff_by_total_lose, 
          total_game_count = sum(count_lose, count_win)) %>% 
arrange(names)

head(data_summary)
```



# Eigenvalues Ranking


```{r}
data_sum = data %>% mutate(
                s_visitor = visitor_score/(visitor_score + home_score),     # calculate s_ij
                s_home = home_score/(visitor_score + home_score)) %>%       # calculate s_ji
   group_by(visitor, home) %>%                                              # group to collapse the multiple games by (visitor vs home)
   summarize(s_visitor = sum(s_visitor) , s_home = sum(s_home)) %>%         # sum to between  multiple games by (visitor vs home)
   mutate(s_visitor = s_visitor/(s_visitor + s_home) , s_home = s_home/(s_visitor + s_home)) # get the a new ration from all multiple games

head(data_sum)

#  Helper function
INDEXOF = function ( s ) {match(s, data_summary$names)}
```



### A Matrix
```{r}


# A Matrix 
A = diag(1, nrow= length(teams))
#  loop through every games
for(i in 1:nrow(data_sum)){
  # Both game (A vs B) and (B vs A) affect the score S_ab and S_ba so we need to keep updating as we see more match between A & B 
  new_sum_s_visitor = A[INDEXOF(data_sum$visitor[i]), INDEXOF(data_sum$home[i])]  + data_sum$s_visitor[i] 
  new_sum_s_home = A[INDEXOF(data_sum$home[i]), INDEXOF(data_sum$visitor[i])] + data_sum$s_home[i]
  total_sum =new_sum_s_home + new_sum_s_visitor
  # update the new value of s_ij and s_ji 
  A[INDEXOF(data_sum$visitor[i]), INDEXOF(data_sum$home[i])] = new_sum_s_visitor/total_sum
  A[INDEXOF(data_sum$home[i]), INDEXOF(data_sum$visitor[i])] = new_sum_s_home/total_sum
}
A
```

### Eigenvalue and Eigenvector

```{r}


A.eigen = eigen(A)
A.lambda = as.numeric(A.eigen$values[1])
A.eigenvector =as.numeric( A.eigen$vectors[,1])

A.lambda
A.eigenvector

```
### Eigen Ranking Result

```{r}



eigen_result =  data.frame(eigen_ranking = A.eigenvector) %>%
  mutate(names = data_summary$names)%>% # add corresponding team name
 arrange(desc(eigen_ranking)) %>%       # reorder team based on ranking score
mutate(eigen_rank = row_number())       # assign rank to teams

eigen_result
```

# Colley Ranking

The `data_summary` data frame contains necessary information sucha as total number of lose games, win games  and total game about each team for creating Colley's matrix.

```{r}


head(data_summary)
```

### N Matrix
```{r}

#  count of game between team i and j 
n_matrix = Matrix(0, nrow = length(data_summary$names), ncol = length(data_summary$names))
#  Populate count
for(i in 1:nrow(data)) {       # for-loop over rows
  team_i = data$visitor[i]     # visitor team name
  team_j = data$home[i]        # home team name
  n_matrix[INDEXOF(team_i), INDEXOF(team_j)] = n_matrix[INDEXOF(team_i), INDEXOF(team_j)] + 1   # count the time both team meets
  n_matrix[INDEXOF(team_j), INDEXOF(team_i)] = n_matrix[INDEXOF(team_i), INDEXOF(team_j)]       # both count are the same for N_ji and N_ji
}
```

### C Matrix
```{r}


# the Colley Matrix 
C_matrix = diag(2 + data_summary$total_game_count, ncol = length(teams) ) - n_matrix

```
#### B vector

```{r}


# The b vector n x 1  on the right with b_i = 1 + 0.5*(wi − li)
colley_b_vector  = 1 + 0.5*(data_summary$count_win - data_summary$count_lose)

colley_b_vector
```

### Colley Ranking Result

```{r}
colley_ranking_vector = solve(C_matrix, colley_b_vector)

colley_result = data.frame(colley_ranking = as.matrix(colley_ranking_vector)) %>%
 mutate(names = data_summary$names) %>%     # add corresponding team names
 arrange(desc(colley_ranking)) %>%          # reorder based on ranking score
mutate(colley_rank = row_number())          # assign ranking

# colley_result
colley_result %>% arrange(names)
```


# Result

### Five Thirty Eight Ranking
Looking at Ranking from the Five Thirty Eight

```{r}



fte_ranking  = data.frame(names = c("Colorado Avalanche","Florida Panthers","Carolina Hurricanes","Tampa Bay Lightning","Toronto Maple Leafs","Minnesota Wild","Calgary Flames","Boston Bruins"	,"New York Rangers"	,"St. Louis Blues","Edmonton Oilers","Pittsburgh Penguins","Vegas Golden Knights","Washington Capitals","Nashville Predators","New York Islanders","Vancouver Canucks"	,"Dallas Stars"		,"Los Angeles Kings","Winnipeg Jets","Columbus Blue Jackets"	,"Ottawa Senators"	,"Buffalo Sabres"	,"Seattle Kraken"	,"Anaheim Ducks"	,"San Jose Sharks"	,"Chicago Blackhawks"	,"Detroit Red Wings"	,"New Jersey Devils"	,"Arizona Coyotes"	,"Montreal Canadiens","Philadelphia Flyers"), fte_rank = 1:32)
```
```{r}



#  the ranking based on two systems listed by team names
ranking = inner_join(eigen_result, colley_result) %>% 
 inner_join(fte_ranking) %>%
 select(-c(eigen_ranking, colley_ranking))

head(ranking)
```
```{r}


plot_data = ranking %>% 
 pivot_longer(cols = c(eigen_rank,colley_rank, fte_rank), names_to = "ranking_type", values_to = "ranking_order") %>%
 mutate(ranking_type = fct_relevel(ranking_type,c("eigen_rank", "colley_rank", 'fte_rank')))
head(plot_data)

```
```{r}
library("ggbump")
```
```{r}
plot_data %>%
 ggplot(aes( x=ranking_type, y = ranking_order, color = as.factor(names), group=names)) + 
 geom_bump(size = 1, smooth = 30, show.legend = T) + 
 geom_label(data = subset(plot_data, ranking_type == "eigen_rank"), aes(x= ranking_type, label = names, y = ranking_order), nudge_x = -0.6, hjust = 0, size =3) + 
 geom_label(data = subset(plot_data, ranking_type == "fte_rank"), aes(x= ranking_type, label = names, y = ranking_order), nudge_x = 0.6, hjust = 1, size =3)  + 
 scale_x_discrete(labels = c("Eigenvector Ranking","Colley's Ranking", "Five Thirty Eight Ranking"), 
 position = "top")+
 scale_y_reverse(breaks = 1:32)+
 labs(title = "Team Ranking", 
      x = element_blank(), 
      y = element_blank()) +
 theme(legend.position = 'none')
```