---
title: "| Computational Linear Algebra    \n| Midterm 1: Take-Home Portion \n"
author: "Vichearith Meas"
output:
  bookdown::html_document2:
    number_sections: false
    split_by: none
    toc: yes
    toc_float: yes
    toc_depth: 2
    theme: cerulean
    css: style.css
---

***Due Tuesday, March 1, 2022, 9:00AM sharp. Please submit an html (or pdf) of your completed work on Moodle. If you wish to write things on paper to complement your typed work, you can upload a picture/scan of your work to Moodle.***

The following line sources functions from the class file `365Functions.r`. Feel free to use any of these functions.
```{r,message=FALSE,warning=FALSE}
source('https://www.macalester.edu/~dshuman1/365/365Functions.r')
```


```{r include=FALSE}
library(Matrix)
library(ggplot2)
library(tidyverse)
library(lemon)
library(knitr)
knit_print.data.frame <- lemon_print
```

## Problem 5: Multivariate Rootfinding via Newton's Method

a) Briefly explain why Algorithm 2 is better than Algorithm 1.

::: answer

In Algorithm 2, we solve for $s = DF(x_{k-1})^{-1}F(x_{k-1})$ first. In each iteration, solving for $s$ can be efficient. $s = DF(x_{k-1})^{-1}F(x_{k-1}) \quad \Rightarrow \quad DF(x_{k-1})s = - F(x_{k-1})$. 

Solving this equation is like solving regular $Ax = b$ assuming $DF(x_{k-1}) = A$ and $b = -F(x_{k-1})$. Solving $Ax= b$ is way more efficient than finding $A^{-1}$ and do dot multiplication.  
:::



b) Write a function that implements Algorithm 2
Then test your function example system given in equation (3), with $x_0=(0.5,0.5)$.

::: answer 

```{r}

vnorm <- function(v,p=2) { 
  if ( p =="I") {
    return(max(abs(v)))
  }
  else {
    return(sum(abs(v)^p)^(1/p))
  }
}


mvnewt <- function(F,DF,x0,maxiter=100,tol=1e-8) {
 #  N: dimention of x 
  N = length(x0)
 #  history to save prediction
  history = spMatrix(N,maxiter + 1)
  history[,1] = x0          #add first prediction
  #  record iteration
  steps=0

  for (j in 2:(maxiter+1)) {
    steps <- steps+1
    prevX = history[,(j-1)]
    s = solve(DF(prevX), -F(prevX))
    history[,j] = prevX + s
    if (vnorm(s,2) <= vnorm(prevX,2)*tol) break
  }
  return(list(x=history[,(steps+1)],iterations=steps,history = history))
}


```

```{r}
DF  = function(x){
  u = x[1]
  v = x[2]
  return(matrix(c(-2*u, 1, 2*u , 2*v), nrow = 2, ncol = 2))
}

F = function(x){
  u = x[1]
  v = x[2]
  return(c(-u^2 + v, u^2 + v^2 - 1))
}

x0 = c(0.5,0.5)
mvnewt_estimation = mvnewt(F, DF, x0, tol=1e-30)

mvnewt_estimation$x
F(mvnewt_estimation$x)

```
:::



c) Find the two roots for the system of equations (4).

::: answer

We have:

$$F(u,v,w) =       \begin{pmatrix}
      u^2 + v^2 + w^2 -4  \\
       sin(2\pi v)- w  \\
       u - v^2
      \end{pmatrix}
\\
DF(u,v,w) = \begin{pmatrix}
      2u & 2v & 2w \\
       0 & 2\pi cos(2\pi v) & -1 \\
       1 & -2v & 0 
      \end{pmatrix}$$


```{r}
Fc = function(x){
  u = x[1]
  v = x[2]
  v2 = v*v
  w = x[3]
  return(c(u^2 + v2 + w^2 - 4, sin(2*pi*v) - w , u - v2))
}
DFc = function(x){
  u = x[1]
  v = x[2]
  w = x[3]
  return(matrix(c(
    2*u, 2*v ,  2*w,
    0 ,  2*pi*cos(2*pi*v), -1, 
    1, -2*v, 0
  ), nrow= 3, ncol= 3 , byrow=TRUE))
}
```

```{r}
initial_guess = list(
  c(1,1,1),
  c(-1,1,1),
  c(0,1,-1),
  c(2,-1,1), 
  c(1,2,1),
  c(-1,2,1),
  c(-1,1,-2),
  c(2,0,1), 
  c(20,10,10),
  c(-20,-10,-10),
  c(20,10,-10),
  c(-20,10,10)
)

for( i in initial_guess ){
xc = mvnewt(Fc, DFc, i)
print(xc$x)
print(Fc(xc$x))
}

```
The estimation for the roots given by the algorithm are $(  1.367119749, -1.169238961, -0.873993046)$ and $( 1.367119749, 1.169238961, 0.873993046)$


:::

## Problem 6: Multivariate Rootfinding without a Jacobian Matrix

a) Write a function that implements Algorithm 2

::: answer

```{r}
mvroot <- function(F,J0,x0,maxiter = 100,tol = 1e-8){
 #  N: dimention of x 
  N = length(x0)
 #  history to save prediction
  history = spMatrix(N,maxiter + 1)
  history[,1] = x0          #add first prediction
  #  record iteration
  steps=0
  # delta 

  for (j in 2:(maxiter+1)) {
    steps <- steps+1
    prevX = history[,(j-1)]
    s = solve(J0, -F(prevX))
    x = prevX + s
    history[,j] = x
    t_s = t(s)
    J0 = J0 + ((F(x) - F(prevX) - J0%*%s)%*%t_s)/(as.vector(t_s%*%s))
    if (vnorm(s,2) <= vnorm(prevX,2)*tol) break
  }
  return(list(x=history[,(steps+1)],iterations=steps,history = history))
}
```

:::

Then test your function example system given in equation (3), with $J_0$ equal to the 2x2 identity matrix and $x_0=(0.5,0.5)$

::: answer 
```{r}
J0 = diag(1,nrow = 2)
x0_6 = c(0.5,0.5)
mvroot_estimation = mvroot(F, J0, x0_6, maxiter = 100)
mvroot_estimation
```
:::

b) Compare the convergence rates for the multivariate Newton method and the Jacobian-free method. Do they converge linearly? superlinearly? quadratically? Support your answer with numerical evidence.


::: answer

Solving for exact root: 
$$
\begin{cases}
-u^2 + v = 0 \\
u^2 + v^2 = 1 
\end{cases}\Rightarrow 
v^2 + v -1 = 0 
\Rightarrow 
\begin{cases}
v = \frac{-1+\sqrt{5}}{2}, u =  \sqrt{\frac{-1+\sqrt{5}}{2}} \\
v = \frac{-1+\sqrt{5}}{2}, u =  - \sqrt{\frac{-1+\sqrt{5}}{2}} \\
\end{cases}
$$
```{r}
(root_1 = c(-sqrt((- 1 + sqrt(5) )/2 ), (-1 + sqrt(5))/2))
(root_2 = c(sqrt((- 1 + sqrt(5) )/2 ), (-1 + sqrt(5))/2))

```
Both of the multivariate Newton method and the Jacobian-free method converge linearly. First, I find the 2-norm distance between the prediction given by each method and the actual root. Then I compute the error ratio $\frac{|e_{k+1}|}{|e_k|^q}$ for  $q \in \{1, 1.5, 2\}$ and then plot the graph. 

The plot show that for both method to be linearly converge. 

```{r message=FALSE, warning=FALSE}
# mvnewt_estimation$history
mvnewt_errors = rep(0, ncol(mvnewt_estimation$history)) 
for (i in 1:ncol(mvnewt_estimation$history)){
  mvnewt_errors[i] = vnorm(mvnewt_estimation$history[,i] - root_2,p = 2)
}

mvroot_errors = rep(0, ncol(mvroot_estimation$history)) 
for (i in 1:ncol(mvroot_estimation$history)){
  mvroot_errors[i] = vnorm(mvroot_estimation$history[,i] - root_1,p = 2)
}
```



```{r message=FALSE, warning=FALSE}
mvn_q = data.frame(x = 1:34, 
                   linear = mvnewt_errors[2:35]/mvnewt_errors[1:34],
                   superlinear = mvnewt_errors[2:35]/(mvnewt_errors[1:34]^1.5),
                  quadratic = mvnewt_errors[2:35]/(mvnewt_errors[1:34]^2))

mvr_q =  data.frame(x= 1:15, 
                    linear = mvroot_errors[2:16]/mvroot_errors[1:15],
                   superlinear = mvroot_errors[2:16]/(mvroot_errors[1:15]^1.5),
                    quadratic = mvroot_errors[2:16]/(mvroot_errors[1:15]^2))
mvr_q

mvn_q

ggplot(mvn_q, aes(x = x)) + 
  geom_line(aes(y = linear, colour = "Linear")) + 
  geom_line(aes(y = quadratic, colour = "Quadratic"))+
  geom_line(aes(y = superlinear, colour = "Super Linear"))  + 
  labs(title= 'Error Ratio of Multivariate Rootfinding via Newton’s Method', 
       x = 'Iteration')  + 
  theme(axis.title.x = element_text("Iteration"), 
        axis.title.y = element_text("Error Ratio"))
ggplot(mvr_q, aes(x = x)) + 
  geom_line(aes(y = linear, colour = "Linear")) + 
  geom_line(aes(y = quadratic, colour = "Quadratic"))+
  geom_line(aes(y = superlinear, colour = "Super Linear"))  + 
  labs(title= 'Error Ratio of Multivariate Rootfinding without a Jacobian Matrix', 
       x = 'Iteration') + 
  theme(axis.title.x = element_text("Iteration"), 
        axis.title.y = element_text("Error Ratio"))

```


However, even though they are both linearly converge, for this particular initial value $x_0$, the  `mvnewt` seems to converge faster than `mvroot`. 

```{r message=FALSE, warning=FALSE}
mvnewt_errors =  data.frame(mvnewt_errors, x = 1:101)

mvnewt_errors %>% 
  ggplot() + 
  geom_line(aes(x = x, y = mvnewt_errors ))

mvroot_errors = data.frame(mvroot_errors, x = 1:101)
mvroot_errors %>% 
  ggplot() + 
  geom_line(aes(x = x, y = mvroot_errors))
```

:::


